Advice on constructing composite ordinal outcomes

### 2025-11-06 1_constructing_ordinal_outcomes.tex
A bunch of work

\title{Constructing composite ordinal outcomes for clinical trials}
\author{Leon Di Stefano + Ordinal
Working Group}
\date{\today}

@@

  Ordinal outcomes are an increasingly popular choice for clinical trials. This
  paper is a guide to techniques for constructing ordinal outcomes out of
  pre-existing components, aimed at clinicians and their statistical
  collaborators. We discuss three broad mechanisms for combining existing
  ordinal outcomes---including binary, continuous, and time to event
  outcomes---into an ordinal composite: overriding (or ``sequential
  composition''), tie-breaking (or ``hierarchical composition''), and what we
  term ``parallel composition''. We illustrate these mechanisms with examples
  from the literature and discuss how constructing fine-grained composite
  outcomes can improve validity and efficiency---but also why coarser outcomes
  might sometimes be preferred in practice. Finally, we connect these general
  methods for combining ordinal outcomes with the ``prioritized outcomes''
  literature (including win-based and desirability of outcome ranking (DOOR)
  methodologies), clearly delineating the task of constructing composite
  ordinal outcomes from the related but distinct tasks of choosing effect
  measures, choosing estimators, and handling censoring.

@@

Ordinal outcomes (or endpoints) rank patients trajectories from better to
worse, but do not treat \emph{differences} in individual patient outcomes as
meaningful. Some ordinal outcomes have a long history in specific disease
areas, for example the modified Rankin scale in acute stroke trials
\citep{broderickEvolutionModifiedRankin2017}, and ordinal outcomes have become
increasingly prominent in clinical resarch following the COVID-19 pandemic when
disease-specific ordinal outcome scales were developed by the WHO and NIAID
\citep{beigelRemdesivirTreatmentCovid192020, marshallMinimalCommonOutcome2020}.
Ordinal outcomes have also been used in a number of adaptive platform trials
\citep{angusREMAPCAPRandomizedEmbedded2020, walkerCodesigningNovelOrdinal2025}.

Two desirable features of outcomes in general are \emph{validity} and
\emph{efficiency}. \emph{Valid} outcomes reflect patient benefit and harm;
\emph{efficient} outcomes are able to discriminate among better- and worse-off
patients with relatively small sample sizes. (Strictly speaking efficiency is a
  property not of outcomes themselves but of estimators of effect measures for
  summarizing differences in outcome distribution---more on this in
\ref{sec:prioritized_outcomes}.)

@@

An example of an outcome which is valid but inefficient might be
death as a binary outcome in a context where deaths
are rare.

An example of an outcome which is efficient but which may be invalid
might be continuously-measured concentrations of a biomarker which is a poor
surrogate for disease progression, or a continuous outcome that
excludes mortality.

@@

\section{Three ways to create composite ordinal outcomes}

@@

\subsection{Increasing efficiency by shrinking the largest category}

@@

\subsection{Consider efficiency within levels of strong predictors}

\subsection{Coarsening to avoid comparing ambiguous levels (validity)}

@@

\subsection{Coarsening for ease of communication}

@@

\section{Hierarchical ordinal outcomes and pairwise comparison methods}

>>>>

\title{Advice on constructing composite ordinal outcomes}
\author{Leon Di Stefano + Ordinal
Working Group}
\date{\today}

@@

  Ordinal outcomes are an increasingly popular choice for clinical trials. This
  paper is a guide to techniques for constructing ordinal outcomes out of
  pre-existing components, aimed at clinicians and their statistical
  collaborators. We discuss three broad mechanisms for combining existing
  ordinal outcomes---including binary, continuous, and time to event
  outcomes---into an ordinal composite: (1) overriding (or ``sequential
  composition''), (2) tie-breaking (or ``hierarchical composition''), and what
  we term (3) ``parallel composition''. We illustrate each of these mechanisms
  with examples from the literature. Next, we discuss how constructing
  fine-grained composite outcomes can improve validity and efficiency---but
  also why adding levels sometimes has very little effect on efficiency.
  Finally, we connect these general methods for combining ordinal outcomes with
  the ``prioritized outcomes'' literature---including win-based and
  desirability of outcome ranking (DOOR) methodologies---clearly delineating
  the task of constructing composite ordinal outcomes from the related but
  distinct tasks of choosing effect measures, choosing estimators, and handling
  censoring.

@@

Ordinal outcomes (or endpoints) rank patients trajectories from better to
worse, but do not treat \emph{differences} in individual patient outcomes as
meaningful. Ordinal outcomes have a long history in specific disease
areas---for example the modified Rankin scale in stroke trials---and have
become increasingly prominent in clinical resarch following the COVID-19
pandemic when disease-specific ordinal outcome scales were developed by the WHO
and NIAID \citep{broderickEvolutionModifiedRankin2017,
beigelRemdesivirTreatmentCovid192020, marshallMinimalCommonOutcome2020}.
Ordinal outcomes have also been used in a number of adaptive platform trials
\citep{angusREMAPCAPRandomizedEmbedded2020, walkerCodesigningNovelOrdinal2025}.

Two desirable features of outcomes in general are \emph{validity} and
\emph{efficiency}. \emph{Valid} outcomes reflect differences in patient benefit
and harm; \emph{efficient} outcomes are able to discriminate among better- and
worse-off patients with relatively small sample sizes. (Each of these is a
function of other factors too.)

Ordinal outcomes have the potential to

@@

@@

\section{Three ways to combine existing ordinal scales}

@@

\subsection{More levels: increasing efficiency by
shrinking the largest category proportion}

@@

\subsection{More levels: Consider efficiency within levels of strong predictors}

\begin{figure}
  \includegraphics[width=7in]{stratified_efficiency_plots.png}
  \caption{\textbf{Overriding (sequential composition) illustrated
    with the PASSPORT outcome} Comparison of the ordinal outcome used
  in PASSPORT (depicted in Figure \ref{fig:passport_outcome})} with a
  coarsened 4-level version, stratified by Australisian Triage Scale category.
  \label{fig:stratified_efficiency}
\end{figure}

\subsection{Fewer levels: avoid comparing ambiguous levels (validity)}

@@

\subsection{Fewer levels: ease of communication}

@@

\section{Composite ordinal outcomes and ``win'' methods}
